---
title: "1969_Draft"
author: "Santonastaso-Winston"
date: "July 30, 2022"
output:
  word_document: default
  html_document: default
---

# Was the 1969 draft lottery fair?
It turned out the selection process was biased towards certain months. The process was corrected for 1970. 
The data and story for the comes from :
Winston, Wayne. *Danalytics Stories: Using Data to make Good Things Happen. * Indianapolis, John Wiley & Sons, Inc., 2021.

```{r}
# set the working directory
setwd('C:/Users/tsant/Documents/Data Science/Data_Stories/Chapter 02')

#import the packages needed
#install.packages('readxl')
library(readxl)
#install.packages('dplyr')
library(dplyr)

```
```{r}
#read in data
#We're reading in one worksheet from a file that contains 3 worksheets.
df <- read_excel("DraftData_TS.xlsx", sheet = "Raw_Data")
```

```{r}
#make sure df is a dataframe:
df <- as.data.frame(df)
df$N70 <- as.integer(df$N70)
```

Now, let's explore the data:

```{r}
#Base R
head(df, 10) #gives the first 10 rows of the data frame
tail(df, 10) #gives the last 10 rows of the data frame
dim(df) #gives the dimensions of the table (#rows    #columns)
str(df) #gives the variables, the class of data, and the first few examples of each

#dplyr
View(df) #opens tab in R studio and displays entire data frame
glimpse(df) #similar to str

```

We need to calculate the averages (day number) of N69, and N70 for each month and make a new table of those average.
We also have to calculate the expected average (day number).

```{r}
mean_69 <- c() #establish an empty list
mons <- 1:12 #create list of numbers to iterate calculation through
for (elem in mons) {
  num <- mean(df[df$Month == elem, 'N69']) #calculate the mean of all the numbers in column 'N69' that match the number of the month #from Month column with the iterable (1 - 12)
  mean_69 <- append(mean_69, num) #append that number to the list we're building
  
} 

#do the same for the N70 column:
mean_70 <- c()
for(elem in mons){
  num <- mean(df[df$Month == elem, 'N70'], na.rm = T) #in this one we have to omit the NAs
  mean_70 <- append(mean_70, num)
}

#last, a vector with expected value:
expected <- rep(((366+1)/2), length(mons))

#Month codes:
Months <- 1:12

#let's create a dataframe:
avgs_df <- data.frame(Months = Months, mean_69 = mean_69, mean_70 = mean_70, expected = expected)
```

And a line graph:

```{r}
#we start by setting up the graph, which not only plots the first set of data but establishes the range of the axes and axis titles
plot(avgs_df$Months, avgs_df$mean_69, type = "l", lty = 1, lwd = 1, col = "black", ylim = c(0,max(avgs_df$mean_69)), main = "Average Calendar Day Number for Each Month", xlab = "Month", ylab = "Average Day Number")
#then, layer on each line with a different color
lines(avgs_df$Months, avgs_df$mean_70, lty = 2, lwd = 1, col = "green")
lines(avgs_df$Months, avgs_df$expected, lty = 3, lwd = 2, col = "blue")
#finally, a legend; note the parallel structure of the codes
legend("bottomright", legend =c("avg in '69", "avg in '70", "expected avg"), col = c("black", "green", "blue"), lty = 1:3, cex = 0.8)

```

We will now use a two-sided t-test to determine if there is a sampling bias between the first and second haves of each year.
Let's prep the data for a two sided t-test.

```{r}
#These commands subset each row from N69 or N70 that match months 1-6 or 7-12
Jan_Jun69 <- subset(df$N69, df$Month %in% c(1:6)) 
Jul_Dec69 <- subset(df$N69, df$Month %in% c(7:12))
Jan_Jun70 <- subset(df$N70, df$Month %in% c(1:6))
Jul_Dec70 <- subset(df$N70, df$Month %in% c(7:12))
```

```{r}
t.test(Jan_Jun69, Jul_Dec69)
t.test(Jan_Jun70, Jul_Dec70)
```

We can see that in 1969, there was a 0.0034% chance that those calendar dates were unbiasedly selected for the calendar date when comparing the 1st half of the year with the 2nd half of the year.

But with the change in procedure, in 1977, there's a 77.9% chance that these dates were chosen at random.
